Experience Report: Haskell in Computational Biology

### 2012-03-10 mrfy_experience_report.tex
Attempt at story & code for search.

We wrote a basic search function in only twenty-two lines of code.
The central logic of this search function is simply:
\begin{smallverbatim}
search' (s1:s2:seeds) oldPop hist age =
  let newPop = mutate oldPop
      score = fst $ minimum newPop
      newHist = score : hist
    in if accept newHist age then
          if terminate newHist age then
            (minimum newPop, newHist)
          else
            search' seeds newPop newHist (age + 1)
       else
         if terminate hist age then
           (minimum oldPop, hist)
         else
           search' seeds oldPop hist (age + 1)
\end{smallverbatim}
As one of the parameters to the top-level \texttt{search} function,
we provide a \texttt{SearchStrategy} data type, which we use to 
define specific search techniques.
We require a \texttt{SearchStrategy} to 
define four functions: \texttt{mutate}, \texttt{accept}, \texttt{terminate}, 
and \texttt{initialize}; we use the Haskell module system to allow sharing of 
some of these functions between different search strategies.
For example, both 
\textit{random mutation hill climb} and \textit{simulated annealing} use the 
same initialize and mutate functions, but very different accept functions.
In 
all, we implemented four search strategies: \textit{random mutation hill 
climb}, \textit{simulated annealing}, \textit{multistart simulated annealing}, 
and \textit{genetic algorithm}.
We found the implementation of the search 
function and search strategies to be pleasant, and the resulting code to be 
simple and elegant.
In an imperative language with object orientation like Ruby 
or Python, we likely would have used a class hierarchy to accomplish this.
We 
feel that this would have required more code, because object orientation is 
meant for encapsulating data as well as functions on those data; here we only 
wanted to encapsulate functions, as the data did not change.

>>>>

These various search heuristics differ in how they approach four subproblems:
how they \textit{initialize} the search with a starting guess or guesses;
how they \textit{mutate} a solution into a new one; how they decide whether
to \textit{accept} a solution over a previous one; how they decide when to
\textit{terminate} the search.
For example, random-mutation hill climbing only accepts better guesses,
while simulated annealing sometimes accepts worse guesses; a genetic algorithm
combines ``parent'' guesses in its mutate function, while random-mutation
hill climbing simply permutes a guess into a new one.

We represent these various search heuristics as a \texttt{SearchStrategy}:
\begin{verbatim}
data SearchStrategy = 
      SS { accept :: SearchParameters -> Seed -> History -> Age -> Bool
         , terminate :: SearchParameters -> History -> Age -> Bool
         , mutate :: SearchParameters -> Seed -> QuerySequence ->
            Scorer -> [BetaStrand] -> [SearchSolution] ->
            [SearchSolution]
         , initialize :: SearchParameters -> Seed ->
            QuerySequence -> [BetaStrand]-> [SearchGuess]
         }
type Scorer = QuerySequence -> [BetaStrand] -> SearchGuess -> SearchSolution
\end{verbatim}
This \texttt{SearchStrategy} is passed (as part of \texttt{SearchParameters})
to our generic search function:
\begin{smallverbatim}
data SearchParameters = SP { strategy :: SearchStrategy
                           , generations :: Int
                           }

search :: QuerySequence -> HMM -> [BetaStrand] -> SearchParameters -> [Seed] -> (SearchSolution, History)
search query hmm betas searchP seeds = search' (tail seeds) initialGuessScore [] 0
  where initialGuessScore = map (score hmm query betas) initialGuess

        initialGuess = (initialize strat) hmm searchP (head seeds) query betas

        strat = strategy searchP
        search' :: [Seed] -> [SearchSolution] -> History -> Age -> (SearchSolution, History)
        search' (s1:s2:seeds) oldPop hist age =
          let newPop = mutate' oldPop
              score = fst $ minimum newPop
              newHist = (score, age) : hist
            in if accept' newHist age then
                  if terminate' newHist age then
                    (minimum newPop, newHist)
                  else
                    search' seeds newPop newHist (age + 1)
               else
                  if terminate' hist age then
                    (minimum oldPop, hist)
                  else
                    search' seeds oldPop hist (age + 1)
            where mutate' = mutate strat searchP s1 query (score hmm) betas
                  terminate' = terminate strat searchP
                  accept' = accept strat searchP s2
\end{smallverbatim}
As one of the parameters to the top-level \texttt{search} function,
we provide a \texttt{SearchStrategy} data type, which we use to 
define specific search techniques.
We require a \texttt{SearchStrategy} to 
define four functions: \texttt{mutate}, \texttt{accept}, \texttt{terminate}, 
and \texttt{initialize}; we use the Haskell module system to allow sharing of 
some of these functions between different search strategies.
For example, both 
\textit{random mutation hill climb} and \textit{simulated annealing} use the 
same initialize and mutate functions, but very different accept functions.
In 
all, we implemented four search strategies: \textit{random mutation hill 
climb}, \textit{simulated annealing}, \textit{multistart simulated annealing}, 
and \textit{genetic algorithm}.
We found the implementation of the search 
function and search strategies to be pleasant, and the resulting code to be 
simple and elegant.
In an imperative language with object orientation like Ruby 
or Python, we likely would have used a class hierarchy to accomplish this.
We 
feel that this would have required more code, because object orientation is 
meant for encapsulating data as well as functions on those data; here we only 
wanted to encapsulate functions, as the data did not change.
